Negative Selection in the 20th Century

### 2020-05-22 why-negative-selection.Rmd
Weight data for parents. standard_ggplot for plots. Attempt (but fail) to run multinomial logits.

theme_set(theme_minimal())
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey50", "TRUE" = "orange2")

@@

my_fill_scale <- function (n, aesthetics = "fill", ...) {
  if (n == 2) {
    scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

@@

```{r plot-means-over-time, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, score, group = score_name)) + 
      geom_line() +
      facet_wrap(~score_name) +
      theme(
        strip.text  = element_text(size = 7), 
        axis.text.x = element_text(size = 7)
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")
```

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      filter(reg.type == "raw") %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = dep.var, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      my_labs() +
      pval_scale() +
      my_fill_scale(n = 2, direction = -1, guide = guide_legend(title = "")) +
      my_theme()

@@

res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2, direction = -1) +
        my_labs() +
        my_theme()

@@

```{r plot-n-partners, fig.cap = "Effect sizes on number of children by lnumber of sexual partners", fig.height = 4}

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
        colour = p.value < 0.05/{{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex)) +
        my_fill_scale(n = 2) +
        pval_scale() +
        my_theme(legend.position = "bottom") +
        my_labs(x = "", y = "")

@@

@@

drake::loadd(res_weighted)
res_weighted_all <- res_weighted %>% 
      left_join(
        res_all %>% 
              filter(dep.var == "N children", reg.type == "raw") %>% 
              select(term, estimate),
        by = "term",
        suffix = c("_weighted", "_unweighted")
      ) %>% 
      mutate(
        prop_wtd_unwtd = estimate_weighted/estimate_unweighted
      )

mean_prop_wtd_unwtd <- mean(res_weighted_all$prop_wtd_unwtd)

@@

As a fix for ascertainment bias in the UK Biobank sample, we weight participants
by sex, age (within the 40-71 age group), and education. Our sample weights are
calibrated based on the 2006 General Household Survey. We then rerun the basic
*children* regressions. Figure \@ref(fig:plot-weighted) shows the results, with
unweighted estimates plotted for comparison. Weighting increases effect sizes on
average by `r 100 * (mean_prop_wtd_unwtd - 1)` per cent.

```{r plot-weighted, fig.cap = "Effects of polygenic scores on number of children, regressions weighted by education levels within age categories"}

n_regs <- as.double(nrow(res_weighted_all))

res_weighted_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate_weighted, order_abs(1))
      ) %>% 
      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
        names_to = "Regression", values_to = "estimate") %>% 
      mutate(
        Regression = sub("estimate_", "", Regression)
      ) %>% 
      ggplot(aes(y = term, x = estimate, shape = Regression,
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
      labs(x = "Effect size", y = "") +
      scale_shape_manual(values = c("weighted" = 22, "unweighted" = 21)) +
      pval_scale(aesthetics = "colour") +
      my_theme()

```{r plot-sibs-weighted, eval = FALSE}

# what this shows is that even after weighting, sibling effect sizes are
# still stronger. In fact weighting hasn't changed things much.
drake::loadd(res_weighted, res_sibs_weighted)
n_regs <- as.double(nrow(res_weighted))
# bind_rows(children = res_weighted, sibs = res_sibs_weighted, .id = "Dep. Var.") %>% 
res_all %>% 
      filter(reg.type == "raw") %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = `dep.var`,
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      labs(x = "Effect size", y = "") +
      my_fill_scale(n = 2) +
      pval_scale(aesthetics = "colour")

@@

These results suggest that previous work may have underestimated the level of
selection occurring in the population. [XXX who?]

@@

A possible answer is given by the economic theory of fertility [XXX cite Becker 1960].
According to this, increases in potential earned income affect fertility via two opposing channels. There is an "income effect" by which children become more
affordable, like any other good. Since childrearing has a cost in time, there is
also a substitution effect: the opportunity cost of childrearing increases if
one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children.

@@

res_age_birth_parents_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1)),
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's age", 
          "moth_age_birth" = "Mother's age"
        )
      ) %>% 
      ggplot(aes(estimate, term, fill = `Dependent variable`, 
        colour = p.value < 0.05 / {{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2, direction = -1) +
        my_labs()

@@

Polygenic scores could capture effects that are really due to population 
stratification, although would not change our results for natural selection of
the scores. In \@ref(fig:plot-regs-controlled) we show results for selection on
polygenic scores residualized for the top 100 principal components of the
genetic data, calculated within the UK Biobank population. (XXX Abdel details.)

@@

```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children.", fig.height = 7}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      facet_wrap(vars(dep.var)) +
      pval_geom_point(alpha = .7) +
      my_labs() +
      pval_scale() +
      my_fill_scale(n = 2, guide = guide_legend(title = ""))

@@

@@

```{r plot-sd-years, eval = FALSE}
# inequality over time. Nothing to see, almost everything is flat.
grid <- expand_grid(score_name = score_names, YOB = 1940:1965)  
grid$sd <- pmap_dbl(grid, ~sd(famhist[famhist$YOB==.y, ][[.x]], na.rm = TRUE))
ggplot(grid, aes(YOB, sd)) + 
      geom_line() + 
      facet_wrap(~score_name, scales = "free_y", ncol = 4)
```

```{r rgs-by-effect-size, eval = FALSE, fig.cap = "Genetic correlations with EA3 by effect size on n siblings."}
rgs <- left_join(rgs, res %>% filter(reg.type == "residualized"), by = c("p2" = "term"))

rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_smooth(method = lm) +
        geom_point(colour = "red2") + 
        geom_label(aes(label = p2, hjust = "left", vjust = "middle"), 
          size = 3, nudge_x = 0.02, alpha = 0.5, 
          label.padding = unit(0.1, "lines")) +
        labs(x = "rG", y = "Effect size", 
          title = "rG with EA3 vs. effect size on # siblings")
```

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(2)),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings"),
        dep.var = fct_relevel(dep.var, "N Siblings")
      ) %>% 
      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
        fill  = PCs)) + 
        pval_geom_point(alpha = 0.8) +
        my_vline +
        facet_wrap(vars(dep.var)) +
        pval_scale() +
        my_fill_scale(n = 2, direction = -1) +
        my_labs() +
        theme(legend.position = "bottom")
```

>>>>

theme_set(theme_minimal())
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "orange2")

standard_ggplot <- function (
        dfr, 
        fill_col, 
        n_regs, 
        ...,
        score_col = quo(term), 
        order_idx = 1, 
        n_cats    = NULL
      ) {
  if (! missing(score_col)) score_col <- enquo(score_col)
  fill_col <- if (! missing(fill_col)) enquo(fill_col) else quo(NULL)
  if (missing(n_cats)) {
    n_cats <- if (missing(fill_col)) 1 else length(unique(pull(dfr, 
                {{fill_col}})))
  }

  n_regs <- as.double(n_regs)

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        ggplot(aes(estimate, {{score_col}}, fill = {{fill_col}}, ...,
          colour = p.value < 0.05/{{n_regs}})) +
          pval_geom_point() +
          my_vline + 
          pval_scale() +
          my_fill_scale(n = n_cats) +
          my_labs() +
          my_theme()
}

@@

my_fill_scale <- function (n, aesthetics = "fill", ...) {
  if (n == 2) {
    scale_colour_manual(values = c("steelblue4", "indianred2"), 
          aesthetics = aesthetics, ...)
    # scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

# from https://github.com/dgrtwo/drlib/blob/master/R/reorder_within.R
# for ordering within facets
reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  new_x <- paste(x, within, sep = sep)
  stats::reorder(new_x, by, FUN = fun)
}

scale_x_reordered <- function(..., sep = "___") {
  reg <- paste0(sep, ".+$")
  ggplot2::scale_x_discrete(labels = function(x) gsub(reg, "", x), ...)
}

scale_y_reordered <- function(..., sep = "___") {
  reg <- paste0(sep, ".+$")
  ggplot2::scale_y_discrete(labels = function(x) gsub(reg, "", x), ...)
}

@@

```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, score, group = score_name)) + 
      geom_line(colour = "steelblue4") +
      facet_wrap(~score_name) +
      my_hline +
      theme(
        strip.text  = element_text(size = 7), 
        axis.text.x = element_text(size = 7)
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")
```

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      filter(reg.type == "raw") %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = dep.var, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      my_labs() +
      pval_scale() +
      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
      my_theme()

@@

res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

@@

```{r plot-n-partners, fig.cap = "Effect sizes on number of children by number of sexual partners", fig.height = 4}

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      ggplot(aes(y = reorder_within(score_name, estimate, sex, fun = order_abs(1)), x = estimate, fill = `N partners`, 
        colour = p.value < 0.05/{{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        my_theme(legend.position = "bottom") +
        my_labs(x = "", y = "") + 
        scale_y_reordered()

@@

Our results are subject to two kinds of ascertainment bias. First, Biobank is
not a random sample of the UK population: for example, it overrepresents
educated people and high income households. Second, sampling parents of
respondents inherently overweights parents who have many children. For instance,
parents of three children will have three times more children as respondents in
UK Biobank than parents of one child, on average. Parents of no children will,
by definition, not be represented at all.

@@

drake::loadd(res_weighted)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_all)
res_weighted_all <- bind_rows(
        "N children" = res_weighted, 
        "N siblings" = res_sibs_parent_weights, 
      .id = "dep.var") 
res_weighted_all$reg.type <- "weighted"
res_weighted_all <- bind_rows(
        res_weighted_all,
        res_all %>% filter(reg.type == "raw") %>% select(-score_name)
      )

res_w_wide <- full_join(
        res_weighted,
        res_sibs_parent_weights,
        by = "term",
        suffix = c(".children", ".sibs")
      ) 
# barbaric but it works:
res_uw_wide <- left_join(
  res_all %>% filter(dep.var == "N children", reg.type == "raw"),
  res_all %>% filter(dep.var == "N siblings", reg.type == "raw"),
  by = "term",
  suffix = c(".children", ".sibs")
)

res_wide <- left_join(res_w_wide, res_uw_wide, by = "term", 
      suffix = c(".wtd", ".uwtd"))

res_wide %<>% mutate(
        prop_children = estimate.children.wtd/estimate.children.uwtd,
        prop_sibs     = estimate.sibs.wtd/estimate.sibs.uwtd
      )

@@

To compensate for ascertainment bias in the UK Biobank sample, we weight participants
by sex, age (within the 40-71 age group), and education. Our sample weights are
calibrated based on the 2006 General Household Survey. We then calculate the 
proportions of respondents within each sibling group size, based on the respondent
weights, and use this to create weights for the sibling regressions.

We rerun the basic regressions for both children and siblings, using the new
weights. Figure \@ref(fig:plot-res-weighted) shows the results. The weightings
are fairly successful at reconciling the overall size of effects from
siblings and children regressions. Tables \@ref(tab:tbl-sibs-weights) and
\@ref(tab:tbl-children-weights) show the largest and smallest changes due to 
weighting in siblings and children regressions.

```{r tbl-sibs-weights}

stars <- function (x) ifelse(x < 0.05/33, "*", "")

top_bottom_5 <- function (x, caption) {
  x %>% 
      slice(c(1:5, nrow(.) - 0:4)) %>% 
      as_hux() %>% 
      insert_row(rep("...", 4), after = 6) %>% 
      set_number_format(-1, -1, 3) %>% 
      set_number_format(-1, 4, 0) %>% 
      set_align(-1, -1, ".") %>% 
      theme_basic() %>% 
      set_font_size(8) %>% 
      set_caption(caption) %>% 
      set_caption_pos("bottom") %>% 
      set_caption_width(1) %>% 
      add_footnote("Insignificant effects not shown", font_size = 7) %>% 
      set_all_padding(0) 
}

res_wide %>% 
      arrange(desc(prop_sibs)) %>% 
      select(term, matches("estimate.sibs"), matches("p.value.sibs"), 
        prop_sibs) %>% 
      filter(p.value.sibs.uwtd < 0.05/33) %>% 
      transmute(
        Score = pretty_names(term),
        `Weighted estimate` = estimate.sibs.wtd,
        `Unweighted estimate` = estimate.sibs.uwtd,
        `Proportion (%)` = prop_sibs * 100
      ) %>% 
      top_bottom_5(caption = "Effects of weighting on sibling regressions: top and bottom 5 changes")
```

```{r tbl-children-weights}

res_wide %>% 
      arrange(desc(prop_children)) %>% 
      select(term, matches("estimate.children"), matches("p.value.children"), 
        prop_children) %>% 
      filter(p.value.children.uwtd < 0.05/33) %>% 
      transmute(
        Score = pretty_names(term),
        `Weighted estimate` = estimate.children.wtd,
        `Unweighted estimate` = estimate.children.uwtd,
        `Proportion (%)` = prop_children * 100
      ) %>% 
      top_bottom_5(caption = "Effects of weighting on children regressions: top and bottom 5 changes")
```

```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Weighted regressions."}

res_weighted_all %>% 
      rename(
        "Dependent variable" = dep.var, 
      ) %>% 
      filter(reg.type == "weighted") %>% 
      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
        shape = reg.type) + 
        scale_shape_manual(values = c(21, 4))

@@

These results suggest that previous work may have underestimated the level of
selection occurring in the population. [TODO: who?]

@@

A possible answer is given by the economic theory of fertility [TODO cite Becker 1960].
According to this, increases in potential earned income affect fertility via two opposing channels. There is an "income effect" by which children become more
affordable, like any other good. Since childrearing has a cost in time, there is
also a substitution effect: the opportunity cost of childrearing increases if
one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children.

@@

res_age_birth_parents_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1)),
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's age", 
          "moth_age_birth" = "Mother's age"
        )
      ) %>% 
      ggplot(aes(estimate, term, fill = `Dependent variable`, 
        colour = p.value < 0.05 / {{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs()

@@

Polygenic scores could capture effects that are really due to population 
stratification, although would not change our results for natural selection of
the scores. In \@ref(fig:plot-regs-controlled) we show results for selection on
polygenic scores residualized for the top 100 principal components of the
genetic data, calculated within the UK Biobank population. (TODO Abdel details.)

@@

```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free_y") +
        pval_geom_point(position = position_dodge(0.4)) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered()

@@

## Genetic correlations with EA3

Another way to examine the "earnings" theory of natural selection is to compare
effect sizes of PGS with their genetic correlation with educational attainment
(EA3). Since EA3 strongly predicts earnings, if earnings drives differences
in fertility, we'd expect a correlation between the two sets of results. Figure
\@ref(fig:plot-rgs-by-effect-size) shows this is so.

TODO: ref source of data.

```{r plot-rgs-by-effect-size, fig.cap = "Effect size on n siblings by genetic correlation with EA3"}

drake::loadd(rgs)
drake::loadd(res_all)

rgs <- left_join(rgs, 
        res_all %>% filter(reg.type == "raw", dep.var == "N children"), 
        by = c("p2" = "term"))

rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_point() +
        my_vline +
        my_hline +
        labs(x = "rG with EA3", y = "Effect on number of children")

@@

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(2)),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings"),
        dep.var = fct_relevel(dep.var, "N Siblings")
      ) %>% 
      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
        fill  = PCs)) + 
        pval_geom_point(alpha = 0.8) +
        my_vline +
        facet_wrap(vars(dep.var)) +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        theme(legend.position = "bottom")
```
